\documentclass[a4paper,USenglish,cleveref,autoref,thm-restate,anonymous]{lipics-v2021}       % onecolumn (second format)

\bibliographystyle{plainurl}% the mandatory bibstyle

\title{Understanding Feedback Mechanisms in Machine Learning Jupyter Notebooks}
%\titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line

%TODO mandatory, please use full name; only 1 author per \author macro; first
%two parameters are mandatory, other parameters can be empty. Please provide at
%least the name of the affiliation and the country. The full address is
%optional. Use additional curly braces to indicate the correct name splitting
%when the last name consists of multiple name parts.
\author{Arumoy Shome}{Delft University of Technology, The Netherlands}{a.shome@tudelft.nl}{}{}

% \author{Joan R. Public\footnote{Optional footnote, e.g. to mark corresponding author}}{Department of Informatics, Dummy College, [optional: Address], Country}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}

% \authorrunning{J. Open Access and J.\,R. Public} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

% \Copyright{Jane Open Access and Joan R. Public} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

% \ccsdesc[100]{\textcolor{red}{Replace ccsdesc macro with valid one}} %TODO mandatory: Please choose ACM 2012 classifications from https://dl.acm.org/ccs/ccs_flat.cfm 

% \keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords

% \category{} %optional, e.g. invited paper

% \relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

%\funding{(Optional) general funding statement \dots}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

% \acknowledgements{I want to thank \dots}%optional

%\nolinenumbers %uncomment to disable line numbering



%Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\EventEditors{John Q. Open and Joan R. Access}
\EventNoEds{2}
\EventLongTitle{42nd Conference on Very Important Topics (CVIT 2016)}
\EventShortTitle{CVIT 2016}
\EventAcronym{CVIT}
\EventYear{2016}
\EventDate{December 24--27, 2016}
\EventLocation{Little Whinging, United Kingdom}
\EventLogo{}
\SeriesVolume{42}
\ArticleNo{23}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{hyperref}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{tabularx}
\usepackage[most]{tcolorbox}
\usepackage{listings}
\usepackage{subcaption}
\usepackage{longtable}
\usepackage{wrapfig}
\usepackage{framed}

\newcommand{\highlight}[1]{
  \begin{framed}%
    \noindent#1
\end{framed}}

\definecolor{backcolor}{rgb}{0.95,0.95,0.92}
\colorlet{punct}{red!60!black}
\definecolor{delim}{RGB}{20,105,176}
\colorlet{numb}{magenta!60!black}

\lstdefinestyle{mystyle}{
  language=Python,
  basicstyle=\ttfamily\footnotesize,
  keywordstyle=\color{delim},
  stringstyle=\color{punct},
  commentstyle=\color{numb},
  breakatwhitespace=true,
  breaklines=true,
  keepspaces=true,
  showspaces=false,
  showstringspaces=false,
  showtabs=false,
  tabsize=2,
  stepnumber=1,
  numbersep=3pt,
  captionpos=b,
  % numbers=left,
  % numberstyle=\tiny\color{gray},
  backgroundcolor=\color{backcolor},
  % aboveskip=0pt,
  % belowskip=0pt
}
\lstset{style=mystyle}

\lstset{emph={%
    assert%
  },emphstyle={\color{delim}}%
}%

\begin{document}


\maketitle

\begin{abstract}
\end{abstract}


\section{Methodology}

The research goal is to \emph{characterize} the validation methods used in ML
Jupyter notebooks. This research goal guides the data collection and analysis
methodology outlined below.

\subsection{Data Collection}
We collect Jupyter notebooks authored in the Python programming language from
two distinct sources, namely GitHub and Kaggle.
We

\subsection{Clustering}

TODO: We remove duplicate statements prior to clustering. We applied clustering
to the entire data and manually analyzed X statements. We observe that the
duplicate statements conflates the clusters. Thus a random sample from the
clusters does not capture diverse. We deliberately trade-off diversity with
representativeness for the taxonomy.

We use CodeBERT to obtain the code embeddings.
This is important since the Python statements in our dataset have semantic meaning.
E.g., \lstinline{assert X == Y} is semantically equivalent to \lstinline{assert X is Y}.

We apply UMAP to reduce the dimensions of the embeddings to 10.

We apply HDBSCAN to cluster the embeddings into distinct groups based
on the similarity of the embeddings.

\subsection{Sampling}

We use Proportional Stratified Sampling to obtain a representative sample of
validation statements.
We sample separately for GitHub and Kaggle.
In total, we obtain X statements from GitHub and Y statements from Kaggle.

We observe validation statements that were exploratory in nature (EC:EXP).
Although they appeared in an ML notebook, the purpose of the statement was not
to validate, but to explore perhaps an API end-point or how a function from an
imported library works.
We discarded such statements.

We observe assert statements what did not validate anything in the pipeline (EC:HW).
But they were written in academic notebooks which served as assignments (e.g., autograder).
We discarded such statements.

We observe code cells where author wrote documentation using Python block comments (EC:ERR).
This produces an output (which is why our data collection script captured them).
However, these are incorrect use of a code cell (should have used markdown cell
instead). We discard such statements.

We discard statements where we cannot gather sufficient context
as the notebooks are not authored in English (EC:ENG).

TODO: cluster 6 contains statements such as \lstinline{plt.ylabel('Win %')}
which has no functional intent relevant to this study (EC:NOINT).
Analyze all statements in such groups, and exclude entirely prior to sampling.
Say that we discovered them during our preliminary analysis.

We observe statements that perform scientific simulations (EC:NOML). These
statements were captured since the notebooks import statistical function
implemented in the libraries we used to identify ML related notebooks. We
discard these statements.

In total, we find X such statements. We follow the same methodology outlined in
the sections above to sample new statements from the respective clusters, until
the quota is filled.

Although several notebooks are assignments or tutorials, the statements are
still valuable and apply to the real-world workflow of developing ML prototypes.

\subsection{Data Analysis}

TODO: Visualizations like \texttt{barplot} or \texttt{catplot} that inspect the
size or count of a feature should be EXP-STRUCT.

TODO: statements associated with train/test splits should go under PREP stage.
The rationale is that it is a data transformation operation.
DATA is about ingestion.
TRAIN is about optimization.
The split is a precondition for training, not a part of it.

TODO: Fine-tune the rationale for each category. There is some ambiguity around
a few e.g., VAL-BOOl versus VAL-EXIST. Also sometime unclear whre visualizations
go.

TODO: Revisit and verify all labels, make sure I didn't mislabel anyhing.

TODO: Revisit the PREP and FEAT stage labels. Consider merging together (I remember
seeing a paper that did that).

TODO: There is some ambiguity between the EXP-INSPECT and EXP-COMPUTE labels.

I think the use of precondition assertions are interesting. E.g., GH227 is a
precondition assertion defined inside a function that creates a visualization.
Writing a test for such a function does not make sense,
yet we might make decisions downstream using this visualization.
In such a case, using a precondition assert that explicitly fails
when there is a fault in the data prevents error being propagated.

\section{Discussion}

Read relevant literature on ML bugs, faults, failures and crashes. Wang 2025 has
a nice table that summarizes prior work. And we are already citing the taxonomy
presented by Morovati 2024.

\section{Threats to Validity}

One threat to validity is that the last cell statement and the output of the
cell may not match or be in sync if say the code in the cell was updated but
never executed.
I wonder if Wang 2025 address this, especially because they looked at error
messages and linked them to the root cause.

\bibliography{bibliography}

\end{document}
